A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-02-01 analysis/paper/paper.Rmd
update paper

@@

## Method quality control

@@

<p class='wip'> From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5% of genes were assigned to be part of the main, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1). 
</p>

@@

Similar to [@schaffter_genenetweaversilicobenchmark_2011], we sample the different parameters from random distributions `r ref("stable", "samplers")`.

@@

Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?. To estimate a cell's position during a simulation within the trajectory structure, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.

## Common normalisation pipeline
We used a standard single-cell RNA-seq preprocessing pipeline which uses the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both all zero genes and all zero cells until convergence.
## Evaluation metrics
## Trajectory types

@@

```{r}
# what commit is this file at? You may need to change the path value
# if your Rmd is not in analysis/paper/
git2r::repository(".")
```
[a]Outdated

>>>>

## Trajectory types

To classify the output of methods and the network `r ref("sfig", "trajectory_type_tree")`.

```{r}
add_sfig({
readRDS(figure_file("trajectory_type_tree.rds", experiment_id="dataset_characterisation/trajectory_types"))
}, 
"trajectory_type_tree", 
"Different trajectory types can be classified in a directed acyclic graph, in which certain types are a generalisation of other trajectory types.", 
10, 
10
)
```

@@

<p class='wip'> We wrapped each method (available at https://www.github.com/dynverse/dynmethods) and postprocessed its output into a common probabilistic trajectory model. This model consists of two parts. The milestone network represents the overall network structure, and contains of edges between different milestones and the length of the edge between them. The milestone percentages contain for each cell its position between two or more milestones, and sums for each cell to one. Linear methods are the easiest to wrap into this format, with two milestones and all cells in between. Methods such as 
</p>## Method quality control

@@

From these module networks we generated gene regulatory networks in two steps: generation of the main regulatory network, and addition of extra target genes from real regulatory networks. For each dataset, we used the same number of genes as were differentially expressed in the real datasets. 5% of genes were assigned to be part of the main, and were randomly distributed among all modules (with at least one gene per module). We sampled edges between these individual genes (according to the module network) using an uniform distribution between 1 and the number of possible targets in each module. To add additional target genes to the network, we assigned every regulator from the network to a real regulator in a real network (from regulatory circuits [@marbach_tissuespecificregulatorycircuits_2016]), and extracted for every regulator a local network around it using personalized pagerank (with damping factor set to 0.1).

@@

Similar to [@schaffter_genenetweaversilicobenchmark_2011], we sample the different parameters from random distributions, given in `r ref("stable", "samplers")`.

@@

<p class='wip'> Because each cellular simulations follows the trajectory it its own speed, knowing the exact position of a cell within the trajectory structure is not straightforward. Furthermore, the speed by which simulated cells made a decision between two or more alternative paths was highly variable sfig?. To estimate a cell's position during a simulation within the trajectory structure, we therefore used the known progression of the modules, given in `r ref("sfig", "modulenets", "c")`,  as a backbone. We smoothed the expression in each simulation using a rolling mean with a window of 50 time steps, and then calculated the average module expression along the simulation. We used dynamic time warping, implemented in the dtw R package [@giorgino_computingvisualizingdynamic_2009; @tormene_matchingincompletetime_2009], with an open end to align a simulation to all possible module progressions, and then picked the alignment which minimised the normalised distance between the simulation and the backbone. In case of cyclical trajectory structures, the number of possible milestones a backbone could progress through was limited to 20.
</p>

## Common normalisation pipeline
We used a standard single-cell RNA-seq preprocessing pipeline which applies parts of the scran and scater Bioconductor packages [@lun_stepbystepworkflowlowlevel_2016]. The advantages of this pipeline is that it works both with and without spike-ins, and includes a harsh cell filtering which looks at abnormalities in library sizes, mitochondrial gene expression, and number of genes expressed using median absolute deviations (set to 3). We required that a gene has to be expressed in at least 5% of the cells, and that it should have an average expression higher than 0.05. Furthermore, we used the pipeline to select the most highly variable genes, using a false discovery rate of 5% and a biological component higher than 0.5. As a final filter, we removed both all zero genes and cells until convergence.

## Evaluation metrics

We used two different metrics to assess the performance of each TI method, one looking at the accuracy of the cellular ordering, and one looking at the accuracy of the detected trajectory structure. The correlation metric first calculates the geodesic distances between all cells within the trajectory structure and than calculates the Spearman's correlation between these two distance matrices. Cells which are positioned closely to each other in both the prediction and the gold standard will therefore have an equally

@@

```{r}
git2r::repository(".")
```
[a]Outdated
